Viscous Erosion in Porous Media

### 2018-03-15 docs/methods/erosion_methods.tex
NM

We will evolve $\theta(\alpha,t)$ in spectral space and so introduce the Fourier series for $\theta$ and $\NLterm$
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha} \, , \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha} \, .
\end{align}
The careful reader will notice that in~\eqref{thetaSeries} we have decomposed $\theta(\alpha)$ into a linear component and a periodic component. In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k,
\qquad k=-N_\iin/2,\ldots,N_\iin/2-1\, .
\end{align*}
These ODEs can be simplified by applying the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right) \, ,
\end{align}
which leads to the coupled system
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1.
\end{align}
Note that the diffusive term from~\eqref{eqn:thetaevo2} does not appear explicitly in~\eqref{eqn:thetaevo3}, but rather is implied by the integrating factor~\eqref{eq:mu}. This is the idea of an exponential integrator. We solve the resulting system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we use the {\em midpoint} RK2 method so as to maximize the smoothing effect of the diffusive term.
\todo[inline]{Last sentence might need to be clarified.  Not sure about
using `so as to'}

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.80 \textwidth]{./figs/01bodseq.pdf}
\caption{A single body eroding in Stokes flow. As the body erodes, it develops a corner at its front and rear stagnation points, while remaining for-aft symmetric due to time-reversibility. Color indicates the vorticity of the surrounding flow, which, when evaluated on the boundary, gives the local shear stress. In this simulation, we use $N_\iin = 1024$ discretization points on the body, $N_\out = 1024$ points on the outer boundary, a time-step of $\Dt = 1E-6$, and smoothing parameters of $\eps = 10/1024$ and $\sigma = 10/1024$.}
\label{01bodseq} 
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024aa;
% Uses 01circ1024.in with epsfac = 10, sigfac = 10, dt = 1e-6, fixarea = 0, fixpdrop = 0

Figure \ref{01bodseq} shows the erosion of a single body, numerically simulated with $N_\iin = 1024$ discretization points and shown at three evenly spaced times. Here, color represents the vorticity of the surrounding flow. When evaluated on the boundary, vorticity corresponds to the shear stress and thus the local erosion rate. As seen in the left-most snapshot, the body is initially circular and has the strongest vorticity above and below it. The flow is for-aft symmetric as a result of the time-reversibility of the Stokes equations, with vorticity vanishing at the body's front and rear stagnation points. 
At the next instance, the body has lost material due to erosion, but remains for-aft symmetric due to the flow symmetry. The greatest material loss has occurred at the body's top and bottom, where shear is strongest. Interestingly, the body has developed well-defined corners at its front and rear stagnation points. This geometric feature is a consequence of the erosion law, which depends on the {\em absolute} shear stress: wherever the stress vanishes, $\atau$ exhibits a corner, which manifests as a corner in the geometry. As shown in the third snapshot, this corner persists as the bodies becomes even more slender and continues to shrink. In this simulation, the body completely vanishes at a final time of $t_f = 1.79E-2$.
\todo[inline]{Is this the notation we want to use for scientific
notation? Could instead use $1.79 \times 10^{-2}$?}

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.8 \textwidth]{./figs/shrink_intface.pdf}
\caption{
Shape and stress evolution: (a) Interfaces of the eroding body from Fig.~\ref{01bodseq} at evenly-spaced time intervals. Color indicates normalized time, $t/t_f$, where $t_f = 1.79E-2$ is the time at which the body completely vanishes. (b) The distribution of absolute shear stress, $\atau$, along each of these interfaces. Normalized arclength, $s/L$, runs counterclockwise from the rear stagnation point. The stress initially varies over the body's surface, but becomes more uniform over time.}
\label{shrink_intface}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024aa;
% Uses 01circ1024.in with epsfac = 10, sigfac = 10, dt = 1e-6, fixarea = 0, fixpdrop = 0

@@

As a final validation, we now aim to verify that our method converges with second-order accuracy in time. To this end, we run the same single-body simulation from Figs.~\ref{01bodseq} and \ref{shrink_intface} (except with a coarser $N_\iin=256$) to a stopping time of $t_s = 1E-2$. This stopping time is 56\% of the time it takes the body to vanish, and was selected so that the body will have eroded substantially but is not so small that accuracy is compromised. We assess error by measuring the difference between simulations of successive resolutions (i.e.~we verify Cauchy convergence). In particular, we measure the $L^2$-difference of the body shapes obtained at the stopping time. At each stage we halve the timestep $\Delta t$. Table \ref{convtab} shows the $L^2$-errors between successive simulations, along with the order of accuracy as obtained by Richardson-extrapolation-type calculations. The table confirms that our method to converges with second-order accuracy. In the table, we also show the CPU time of the simulation run on a single processor (though the algorithm is easily parallelized), showing a linear increase in computational time.

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Convergence test: We run single-body erosion simulations to a stopping time of $t_s =$ 1E-2 (56\% of the vanishing time) and measure the $L^2$-difference in shapes obtained with successive temporal resolutions. The tests confirm the method to be second-order accurate in time. The last column shows the measured CPU time which increases linearly as the time-step is refined.
}
\vspace{0.3 pc}
\label{convtab}
\begin{tabular}{l l l l}
\hline
\hspace{0.0pc} $\Delta t/t_s$
\hspace{0.5pc} & error 
\hspace{0.5pc} & order
\hspace{0.5pc} & CPU time \\
\hline
%
1/50		& 1.68E-5		& --		& 53 secs     	\\
1/100	& 4.84E-6		& 1.79	& 1.8 mins   	\\
1/200	& 1.30E-6		& 1.90	& 3.5 mins  	\\
1/400	& 3.36E-7		& 1.95	& 6.9 mins  	\\
1/800	& 8.54E-8		& 1.98	& 14 mins   	\\
1/1600	& 2.15E-8		& 1.99	& 28 mins  	\\
1/3200	& 5.40E-9		& 1.99	& 56 mins    	\\
1/6400	& --			& --		& 111 mins	\\
%
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Parameters in this test are: 01circ256, nits = 7, dt0 = 1e-4, tfin = 1e-2
% epsfac = 10, sigfac = 10, fixarea = 0, fixprdop = 0

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Convergence test for multiple-body erosion with fixed-pressure-drop condition. In this test, we use 20 bodies with $N_{\iin} = 128$ grid points on each. We run the simulations to a stopping time of $t_s = 4E-3$ (24\% of the time it takes for {\em all} bodies to vanish). We measure the {\em combined} $L^2$-difference in the shapes of {\em all} bodies. The table confirms that the method maintains second-order accuracy in time, even with multiple bodies and a time-varying inflow rate $\umax$.
%We compare the shapes of {\em all} bodies from simulations of successive resolution using {\combined} $L^2$-norm
}
\vspace{0.3 pc}
\label{convtab}
\begin{tabular}{l l l l}
\hline
\hspace{0.0pc} $\Delta t/t_s$
\hspace{0.5pc} & error 
\hspace{0.5pc} & order
\hspace{0.5pc} & CPU time \\
\hline
%
1/10     	& 2.67E-5  	& --        	& 14 mins  	\\
1/20     	& 7.23E-6  	& 1.88 	& 27 mins  	\\
1/40     	& 1.95E-6  	& 1.89 	& 52 mins  	\\
1/80     	& 4.90E-7  	& 1.99 	& 1.7 hours	\\
1/160     	& 1.25E-7  	& 1.97  	& 3.4 hours	\\
%
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Parameters in this test are: 20circ128, nits = 5, dt0 = 4e-4, tfin = 4e-3
% epsfac = 10, sigfac = 10, fixarea = 0, fixprdop = 1

>>>>

We will evolve $\theta(\alpha,t)$ in spectral space and so introduce the Fourier series for $\theta$ and $\NLterm$
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha} \, , \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha} \, .
\end{align}
The careful reader will notice that in~\eqref{thetaSeries} we have decomposed $\theta(\alpha)$ into a linear component and a periodic component. In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k,
\qquad k=-N_\iin/2,\ldots,N_\iin/2-1\, .
\end{align*}
These ODEs can be simplified by applying the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right) \, ,
\end{align}
which leads to the coupled system
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1.
\end{align}
Note that the diffusive term from~\eqref{eqn:thetaevo2} does not appear explicitly in~\eqref{eqn:thetaevo3}, but rather is implied by the integrating factor~\eqref{eq:mu}. This is the idea of an exponential integrator. We solve the resulting system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we use the {\em midpoint} RK2 method because it maximizes the smoothing effect of the diffusive term.
\todo[inline]{Last sentence might need to be clarified.  Not sure about
using `so as to'}

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.80 \textwidth]{./figs/01bodseq.pdf}
\caption{A single body eroding in Stokes flow. As the body erodes, it develops a corner at its front and rear stagnation points, while remaining for-aft symmetric due to time-reversibility. Color indicates the vorticity of the surrounding flow, which, when evaluated on the boundary, gives the local shear stress. In this simulation, we use $N_\iin = 1024$ discretization points on the body, $N_\out = 1024$ points on the outer boundary, a time-step of $\Dt = 10^{-6}$, and smoothing parameters of $\eps = 10/1024$ and $\sigma = 10/1024$.}
\label{01bodseq} 
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024aa;
% Uses 01circ1024.in with epsfac = 10, sigfac = 10, dt = 1e-6, fixarea = 0, fixpdrop = 0

Figure \ref{01bodseq} shows the erosion of a single body, numerically simulated with $N_\iin = 1024$ discretization points and shown at three evenly spaced times. Here, color represents the vorticity of the surrounding flow. When evaluated on the boundary, vorticity corresponds to the shear stress and thus the local erosion rate. As seen in the left-most snapshot, the body is initially circular and has the strongest vorticity above and below it. The flow is for-aft symmetric as a result of the time-reversibility of the Stokes equations, with vorticity vanishing at the body's front and rear stagnation points. 
At the next instance, the body has lost material due to erosion, but remains for-aft symmetric due to the flow symmetry. The greatest material loss has occurred at the body's top and bottom, where shear is strongest. Interestingly, the body has developed well-defined corners at its front and rear stagnation points. This geometric feature is a consequence of the erosion law, which depends on the {\em absolute} shear stress: wherever the stress vanishes, $\atau$ exhibits a corner, which manifests as a corner in the geometry. As shown in the third snapshot, this corner persists as the bodies becomes even more slender and continues to shrink. In this simulation, the body completely vanishes at a final time of $t_f = 1.79 \times 10^{-2}$.
\todo[inline]{Is this the notation we want to use for scientific
notation? Could instead use $1.79 \times 10^{-2}$? \\
Fine by me (Nick). What about for tables (e.g. table 4)? For tables, I am more used to seeing the 1.79E-2 notation. How about you?}

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.8 \textwidth]{./figs/shrink_intface.pdf}
\caption{
Shape and stress evolution: (a) Interfaces of the eroding body from Fig.~\ref{01bodseq} at evenly-spaced time intervals. Color indicates normalized time, $t/t_f$, where $t_f = 1.79 \times 10^{-2}$ is the time at which the body completely vanishes. (b) The distribution of absolute shear stress, $\atau$, along each of these interfaces. Normalized arclength, $s/L$, runs counterclockwise from the rear stagnation point. The stress initially varies over the body's surface, but becomes more uniform over time.}
\label{shrink_intface}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024aa;
% Uses 01circ1024.in with epsfac = 10, sigfac = 10, dt = 1e-6, fixarea = 0, fixpdrop = 0

@@

As a final validation, we now aim to verify that our method converges with second-order accuracy in time. To this end, we run the same single-body simulation from Figs.~\ref{01bodseq} and \ref{shrink_intface} (except with a coarser $N_\iin=256$) to a stopping time of $t_s = 10^{-2}$. This stopping time is 56\% of the time it takes the body to vanish, and was selected so that the body will have eroded substantially but is not so small that accuracy is compromised. We assess error by measuring the difference between simulations of successive resolutions (i.e.~we verify Cauchy convergence). In particular, we measure the $L^2$-difference of the body shapes obtained at the stopping time. At each stage we halve the timestep $\Delta t$. Table \ref{convtab} shows the $L^2$-errors between successive simulations, along with the order of accuracy as obtained by Richardson-extrapolation-type calculations. The table confirms that our method to converges with second-order accuracy. In the table, we also show the CPU time of the simulation run on a single processor (though the algorithm is easily parallelized), showing a linear increase in computational time.

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Convergence test: We run single-body erosion simulations to a stopping time of $t_s = 10^{-2}$ (56\% of the vanishing time) and measure the $L^2$-difference in shapes obtained with successive temporal resolutions. The tests confirm the method to be second-order accurate in time. The last column shows the measured CPU time which increases linearly as the time-step is refined.
}
\vspace{0.3 pc}
\label{convtab}
\begin{tabular}{l l l l}
\hline
\hspace{0.0pc} $\Delta t/t_s$
\hspace{0.5pc} & error 
\hspace{0.5pc} & order
\hspace{0.5pc} & CPU time \\
\hline
%
1/50		& 1.68E-5		& --		& 53 secs     	\\
1/100	& 4.84E-6		& 1.79	& 1.8 mins   	\\
1/200	& 1.30E-6		& 1.90	& 3.5 mins  	\\
1/400	& 3.36E-7		& 1.95	& 6.9 mins  	\\
1/800	& 8.54E-8		& 1.98	& 14 mins   	\\
1/1600	& 2.15E-8		& 1.99	& 28 mins  	\\
1/3200	& 5.40E-9		& 1.99	& 56 mins    	\\
1/6400	& --			& --		& 111 mins	\\
%
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Parameters in this test are: 01circ256, nits = 7, dt0 = 1e-4, tfin = 1e-2
% epsfac = 10, sigfac = 10, fixarea = 0, fixprdop = 0

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Convergence test for multiple-body erosion with fixed-pressure-drop condition. In this test, we use 20 bodies with $N_{\iin} = 128$ grid points on each. We run the simulations to a stopping time of $t_s = 4 \times 10^{-3}$ (24\% of the time it takes for {\em all} bodies to vanish). We measure the {\em combined} $L^2$-difference in the shapes of {\em all} bodies. The table confirms that the method maintains second-order accuracy in time, even with multiple bodies and a time-varying inflow rate $\umax$.
%We compare the shapes of {\em all} bodies from simulations of successive resolution using {\combined} $L^2$-norm
}
\vspace{0.3 pc}
\label{convtab}
\begin{tabular}{l l l l}
\hline
\hspace{0.0pc} $\Delta t/t_s$
\hspace{0.5pc} & error 
\hspace{0.5pc} & order
\hspace{0.5pc} & CPU time \\
\hline
%
1/10     	& 2.67E-5  	& --        	& 14 mins  	\\
1/20     	& 7.23E-6  	& 1.88 	& 27 mins  	\\
1/40     	& 1.95E-6  	& 1.89 	& 52 mins  	\\
1/80     	& 4.90E-7  	& 1.99 	& 1.7 hours	\\
1/160     	& 1.25E-7  	& 1.97  	& 3.4 hours	\\
%
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Parameters in this test are: 20circ128, nits = 5, dt0 = 4e-4, tfin = 4e-3
% epsfac = 10, sigfac = 10, fixarea = 0, fixprdop = 1
